\RequirePackage[l2tabu, orthodox]{nag}	% tells out-of-date packages
\documentclass[10pt, letterpaper]{scrartcl}

\usepackage{amssymb,amsfonts}
\usepackage{amsthm} % if using theorems and proofs
\usepackage[cmex10]{amsmath}
\usepackage[usenames,dvipsnames,svgnames]{xcolor} % use like \textcolor{red}{..} or {\color{red} ...}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}% can also add 'nohead', or leave blank

%\usepackage{url}

\usepackage{cancel} % for canceling terms

\usepackage{booktabs}

\usepackage{pdfpages} % for including pdfs


\usepackage{algpseudocode}  %algorithmicx
\usepackage{wrapfig}

%\usepackage{listings}
\usepackage[numbered,framed]{matlab-prettifier}
\lstdefinestyle{mat}{
    frame=single,
    language=matlab,
    showstringspaces=false,
    %  keywordstyle=\bfseries\color{blue},
    %  commentstyle=\color{green},
    %  stringstyle=\color{magenta},
}
\lstset{
    style              = Matlab-editor,
    basicstyle         = \mlttfamily,
    escapechar         = ",
    mlshowsectionrules = true,
}
%% use withL: \lstinputlisting{myODEfunction.m}


\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern} % important!

\usepackage{hyperref}
\hypersetup{pdfauthor={Stephen Becker},
    pdftitle={},
    colorlinks=true,
    citecolor=MidnightBlue,
    urlcolor=Bittersweet,
}
% -- Input macros and such --
\IfFileExists{custom_headers.tex}{
\input{custom_headers}}{
\input{../custom_headers}}{
}
\let\oldenumerate\enumerate
\renewcommand{\enumerate}{
  \oldenumerate
  \setlength{\itemsep}{1em}
  \setlength{\parskip}{1em}
  \setlength{\parsep}{0pt}
}
\usepackage{enumitem}
%\setlist{noitemsep}
% See http://mirrors.fe.up.pt/pub/CTAN/macros/latex/contrib/enumitem/enumitem.pdf
% and now, set par indent to zero
\newlength{\savedparindent}
\setlength{\savedparindent}{\parindent}
\setlist[enumerate]{listparindent=\savedparindent}
%\setlength{\parindent}{0pt} % Jan 2017, HW 3/4, comment this out.

\usepackage{xspace}
%\newcommand{\collaboration}{\textbf{Collaboration Allowed}\xspace}
%\newcommand{\nocollaboration}{\textbf{No Collaboration}\xspace}

\renewcommand{\T}{\mathbb{T}}
\newcommand{\fn}{\widehat{f}_n} % Fourier coefficients
\newcommand{\spi}{\frac{1}{\sqrt{2\pi}}}
\renewcommand{\H}{\mathcal{H}}
\renewcommand{\phi}{\varphi}
\DeclareMathOperator{\ran}{ran} % ker already defined, and with lower case

\newcommand*\Laplace{\mathop{}\!\mathbin\bigtriangleup}
\renewcommand{\SS}{\mathcal{S}} % Schwartz space

% My solution environment
%%\newenvironment{solution}{\setlength{\parindent}{\savedparindent}{\bfseries Solution}:}{}
\makeatletter
\newcommand\solParagraph{\@startsection{paragraph}{4}{\z@}%
%    {-3.25ex \@plus -1ex \@minus -0.2ex}%
    {-.55ex \@plus -1ex \@minus -0.2ex}%
    {0.01pt}%
    {\raggedsection\normalfont\sectfont\nobreak\size@paragraph}%
}
\makeatother
\newenvironment{solution}{\setlength{\parindent}{\savedparindent}\solParagraph{Solution:}}{}

%\newenvironment{solution}{\emph{Solution}:}{}
\newenvironment{instructions}{}{}
\newenvironment{SolnComment}{}{}
\usepackage{comment} 

%\def\solutions{1}  % define solutions  % !!! COMMENT or UNCOMMENT THIS LINE

% NOTE: \begin{solution} should have NO SPACES BEFORE OR AFTER IT,
% 	Can give very weird errors, hiding text, etc. See http://tex.stackexchange.com/a/91431/4603
\ifdefined\solutions
    \newcommand{\solTitle}[1]{#1}
    \excludecomment{instructions}
\else
   \excludecomment{solution}% uncomment this line to hide solution
   \newcommand{\solTitle}[1]{}
   \excludecomment{SolnComment}
\fi


\title{Homework 8 \solTitle{Selected Solutions} \\APPM 4720/5720 Spring 2019 \\ Randomized Algorithms}
%\author{Stephen Becker}
\date{}
\begin{document}
\maketitle
\vspace{-6em}
\textbf{Due date}: Friday, Mar 8 2019

Theme: Entry-wise sampling   \hfill Instructor: Stephen Becker %Dr.\ Becker
%\vspace{1em}

\begin{instructions}
\paragraph{Instructions}
Collaboration with your fellow students is allowed and in fact recommended, although direct copying is not allowed.  Please write down the names of the students that you worked with. The internet is allowed for basic tasks only, not for directly looking for solutions.

An arbitrary subset of these questions will be graded.

%\paragraph{Reading} 
%Read the rest of chapters 9--11 in [BV2004]. 
%%Read chapter 3.1, 3.2 and 3.3 in [BV2004].
%%Students are \textbf{strongly advised} to skim appendices A and C in [BV2004] to look for unfamiliar material (and read in more detail if there is unfamiliar material)
%
%%\vspace{1em}
%%\textbf{Special instructions this week}:
%
\end{instructions}


\vspace{2ex}
In this homework, we will consider an entry-wise sampling scheme. For a given $m\times n$ target matrix $X$, 
consider a sparse matrix $\hat{X}$ where each entry of $\hat{X}$ is chosen independently, and
\begin{equation} \label{eq:1}
\hat{X}_{i,j} = \begin{cases}  X_{i,j}/p_{i,j} & \text{with probability } p_{i,j} \\
0 & \text{with probability } 1- p_{i,j} \end{cases}
\quad\forall\, i=1,\ldots,M, \; j=1,\ldots,N.
\end{equation}

We will use $X$ to be the matrix from the MNIST data we used previously, so $X$  is a $784\times  3000$ matrix.

\begin{enumerate}[align=left, leftmargin=*, label=\sffamily\bfseries Problem \arabic*:]   
 
    \item \ [READING] Skim the paper \href{https://pdfs.semanticscholar.org/aa64/b8fb3382e42f90ee93a1dd0c78f13833963f.pdf}{``Matrix Entry-wise Sampling: Simple is Best''} (Achlioptas, Karnin, and Liberty, KDD 2013).
    
    Compute the quantities involved in equation (9) of that paper for the matrix $X$ from the MNIST data mentioned above.    

    
    \textbf{Deliverable}: the left-hand and right-side quantities in equation (9) of the paper, and a sentence describing if you expect $\ell_1$ or $\ell_2$ sampling to do better (based on equation (9)).

    \item \ [MATH] 
    For $\hat{X}$ sampled according to the above scheme, 
    \begin{enumerate}
        \item Prove $\E[ \hat{X} ] = X$
        \item Calculate $\E[ \hat{X}\hat{X}^T ]$, and discuss how we could make unbiased estimator of $XX^T$. (This would be usesful to estimate covariance matrices).
    \end{enumerate}
    
    
    \item \ [CODING] 
    
%\begin{lstlisting}
%\end{lstlisting}

    Sample $\hat{X}$ as discussed above, according to two schemes:
    \begin{enumerate}
        \item $\ell_1$ sampling: take $p_{i,j} \propto |X_{i,j}|$, e.g.,
        $p_{i,j} = \min\left( 1, c |X_{i,j}|/( \sum_{i',j'}|X_{i',j'}| ) \right)$ and adjust the constant $c$ to vary the expected number of sampled entries.
        \item $\ell_2$ sampling: take $p_{i,j} \propto X_{i,j}^2$
    \end{enumerate}
    
        \textbf{Deliverable}: for several sparsities (i.e., different values of $c$), record the error (in either spectral or Frobenius norm) of $\|X-\hat{X}\|/\|X\|$ for both $\ell_1$ and $\ell_2$ sampling schemes, and make a plot of sparsity vs.\ error. Discuss if the results agree with your prediction based on Problem 1.
        
        Hint: to make sure your code is correct, you might want to run some experiments to verify $\E[ \hat{X} ] = X$, e.g., by looking at $\| 1/N\sum_{k=1}^N \hat{X}_k - X\|$ for independent draws $\hat{X}_k$, and plotting this error on a log-log plot to make sure it decays to $0$.
        
        Bonus (not required, only for fun): try $\ell_2$ sampling with a truncation scheme for small entries. Does this help?  Or, try $\ell_1$ or $\ell_2$ sampling where the normalization is done per row or per column. Does this improve overall errors vs.\ compression?
        
        Double bonus (not required, only for fun): under either sampling scheme, can you make an estimator that is an unbiased estimator of the covariance $1/nXX^T$? You may need to record a little more information than just $\hat{X}$.
        
    
    
    

\end{enumerate}   
\end{document}
